---
title: "Writing Custom ggplot2 geoms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing Custom ggplot2 geoms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains information on how to make a custom `geom` for a ggplot2 object.

```{r setup}
library(ggextend)
library(ggplot2)
```

```{r, sim-data}
dat <- sim_data(t = 100, b = 10, d = 5)
```

With the simple stochastic time-series a ggplot can be produced:

```{r, plot-data}
ggplot2::ggplot(data = dat) + 
  ggplot2::geom_point(mapping = ggplot2::aes(x = time, y = system))
```

We can use a plotting method to automate create the plot. This works as the 
output of `sim_data()` is tagged with the `<bdts>` S3 class and the plotting
method is written to expect this class (`plot.bdts()`).

```{r, plot-method}
plot(dat)
```

We can simulate with 2 rate regimes. The first rate regime has birth and death 
of 1 (i.e. equal probability at each time step of a birth or a death). The second
rate regime has a birth rate of 10 thus the probability of a birth event will be
10 times more likely than a death event.

```{r, sim-data-2-rate}
dat <- sim_data(t = 100, b = 1, d = 1, t2 = 50, b2 = 10)
```

```{r, plot-2-rate}
plot(dat)
```

To visualise where the rate regime changes we can customise our plots with 
shading and annotations to show where the rate regime shift happens.

```{r, plot-2-rate-shade}
plot(dat) + 
  ggplot2::annotate(
    "rect", 
    xmin = min(which(dat$rate == 2)), 
    xmax = max(which(dat$rate == 2)), 
    ymin = min(dat$system), 
    ymax = max(dat$system), 
    alpha = 0.2, 
    fill = "purple"
  ) + ggplot2::annotate(
    geom = "text", 
    label = "Rate Regime 2",
    x = mean(which(dat$rate == 2)), 
    y = mean(dat$system), 
  )
```

This chunk of custom annotation can also be automated by creating a custom ggplot2
`geom`.

## Creating a custom `geom`

It states in the `Geom` documentation (`?ggplot2::Geom`) and [aaa-.R file in the ggplot2 package](https://github.com/tidyverse/ggplot2/blob/6dd9bb66116ea5531c5b649c02f7390387028087/R/aaa-.R) that to write a custom `geom` `ggplot2::Geom` will need to be extended.

Each `geom` is usuall structured to have one function that is called in composed a plot, e.g., `geom_rect()`, and one function that constructs a `<ggproto>` object, e.g. `GeomRect`.

`{ggplot2}` uses the `{grid}` package for create plots. An introduction to the `{grid}` package can be found by running `vignette("grid", package = "grid")` in R.

Example of `Geom` function:

```{r}
GeomRateShift <- ggplot2::ggproto(
  "Geomrateshift", ggplot2::Geom,
  default_aes = ggplot2::aes(colour = NA, fill = "purple", linewidth = 0.5, linetype = 1,
                    alpha = NA),
  required_aes = c("xmin", "xmax", "ymin", "ymax"),
  draw_key = ggplot2::draw_key_polygon,
  draw_panel = function(self, data, panel_params, coord, lineend = "butt", linejoin = "mitre") {
    coords <- coord$transform(data, panel_params)
    coords$alpha <- 0.5
    grob <- grid::rectGrob(
      coords$xmin, coords$ymax,
      width = coords$xmax - coords$xmin,
      height = coords$ymax - coords$ymin,
      default.units = "native",
      just = c("left", "top"),
      gp = grid::gpar(
        col = coords$colour,
        fill = ggplot2::alpha(coords$fill, coords$alpha),
        lwd = coords$linewidth * .pt,
        lty = coords$linetype,
        linejoin = linejoin,
        lineend = lineend
      )
    )
    grob$name <- grid::grobName(
      grob = grob,
      prefix = "geom_rect"
    )
    grob
  }
)
```


After we've written the new `Geom` function, we can write a `geom_*()` function to
call it when building a plot with `{ggplot2}`:

```{r}
geom_rateshift <- function(mapping = NULL, data = NULL,
                           stat = "identity", position = "identity",
                           ...,
                           linejoin = "mitre",
                           na.rm = FALSE,
                           show.legend = NA,
                           inherit.aes = TRUE) {
  ggplot2::layer(
    geom = GeomRateShift,
    data = data,
    mapping = mapping,
    stat = stat,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      linejoin = linejoin,
      na.rm = na.rm,
      ...
    )
  )
}
```

When calling `ggplot2::layer()` in this function the specified `Geom` constructor is now our new custom `GeomRateShift`.

